---
title: "Class11"
author: "Ethan Lai"
date: "2/23/2022"
output: pdf_document
---
Let's call library(DESeq2) and load our two data files
```{r}
library("DESeq2");
metadata<- read.csv(("airway_metadata.csv"))
counts<- read.csv("airway_scaledcounts.csv", row.names=1)
head(counts)
```
There are `r nrow(counts)` rows, ie "genes" in this dataset


```{r}
metadata
```
It looks like we have four treated (ie with drug) and four control (ie no drug).
 Does the drug do anything?
 
 We want to compare treated vs control.
 
 Fist, let's make sure the metadata matches the counts data order
```{r}
all(metadata$id ==colnames(counts))
```
 Let's get going!

First, let's take a summary statistic of all controls vs all treated:
```{r}
#metadata[metadata$dex=="control", "id"]

controls<- counts[metadata$dex=="control"]
treateds<- counts[metadata$dex=="treated"]
head(controls)
head(treateds)

```

Now that we have our control and treated data separated, we'll find the mean count values for each row ie gene. We could use `apply()` or more simply `rowMeans()`

```{r}
control.means<- rowMeans(controls)
treated.means<-rowMeans(treateds)
```

Let's plot control vs treated means:

```{r}
plot( treated.means, control.means,)
```

Most of the data is in the bottom left. Let's make it a log-log plot

```{r}
plot( treated.means, control.means, log="xy")

```
We often use log transformation because it is more intuitive:
```{r}
log2(20/20)
```
Zero fold change

```{r}
log2(40/20)
```
1 fold log change, meaning doubling.


Let's cvompute the log2 fold change

```{r}
log2fc <- log2(treated.means/control.means)
```

Let's make a data.frame to store our results to data:
```{r}
meancounts<-data.frame(control.means, treated.means, log2fc)
head(meancounts)
```
We need a way to remove NaN and -Inf: We use the which() function in a complicated one-liner.
```{r}
to.rm<- unique(which(meancounts[,1:2]==0, arr.ind=TRUE)[,"row"])
mycounts<- meancounts[-to.rm,]
```

Q1. How many genes do we have left?
Q2. How many have a log2fc more than 2?

```{r}
remainingGenes<- nrow(mycounts)
```

There are `r remainingGenes` genes left.
```{r}
greater<- sum(mycounts$log2fc>2)
```
There are `r greater` genes with log2fc >2


# DESEq2

```{r}
library("DESeq2")
```

First we need to setup the object that DEseq needs with 
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

```{r}
dds<- DESeq(dds)
res<- results(dds)
res
```


## A main fesult figure

A common main result figure from this type of analysis is a volcano plot. THis is a plot of log2 fold change vs P value

```{r}

plot(res$log2FoldChange, -log(res$padj))

```

Let's color code our significant hits

```{r}
plot(res$log2FoldChange, -log(res$padj), col=ifelse(res$padj <= 0.05 & (res$log2FoldChange>2 | res$log2FoldChange<=-2), "red", "black"))

```


#Pathway Analysis

Let's explore the GAGE package
```{r}
library(pathview)
library(gage)
library(gageData)
```

```{r}
data(kegg.sets.hs)

# Examine the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 2)

```
Let's store the fold change data:

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```


Run the gage pathway analysis:

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
head(keggres$less, 3)
```

We use the `pathview()` function (in the pathview package)to make a color coded pathway of our RNA seq results

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa00232")

```

