---
title: "class14"
author: "Ethan Lai"
date: "3/4/2022"
output: html_document
---

```{r}
vax <- read.csv( "covid19vaccinesbyzipcode_test.csv" )
head(vax)
```
Q1. What column details the total number of people fully vaccinated?

Column 9, persons_fully_vaccinated

Q2. What column details the Zip code tabulation area?

Column 2, zip_code_tabulation_area

Q3. What is the earliest date in this dataset?

```{r}
min(vax$as_of_date)
```

2021-01-05

Q4. What is the latest date in this dataset

```{r}
max(vax$as_of_date)
```

2022-03-01




As we have done previously, let’s call the skim() function from the skimr package to get a quick overview of this dataset:

```{r}
library("skimr")
skimr::skim(vax)
```

Q5. How many numeric columns are in this dataset?

 10
 
 
Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

18338

Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

0.17

Q8. [Optional]: Why might this data be missing?

Data is redacted for legal reasons


```{r}

library(lubridate)
today()
vax$as_of_date <- ymd(vax$as_of_date)

```

Now we can do math with dates. For example: How many days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]

```
Using the last and the first date value we can now determine how many days the dataset span?

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]

```


Q9. How many days have passed since the last update of the dataset?
```{r}
today() - vax$as_of_date[nrow(vax)]

```
Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

#Working with ZIP codes

One of the numeric columns in the dataset (namely vax$zip_code_tabulation_area) are actually ZIP codes - a postal code used by the United States Postal Service (USPS). In R we can use the zipcodeR package to make working with these codes easier. For example, let’s install and then load up this package and to find the centroid of the La Jolla 92037 (i.e. UC San Diego) ZIP code 

```{r}
library(zipcodeR)
geocode_zip('92037')
```
Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')

```
More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:
```{r}
reverse_zipcode(c('92037', "92109") )

```

#Focus on the San Diego area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. Using dplyr the code would look like this:

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```
Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
byZip <- rowsum(sd$age12_plus_population,sd$zip_code_tabulation_area)
byZip[which.max(byZip),]
```

Zip code 92154, with population 4658277


Using dplyr select all San Diego “county” entries on “as_of_date” “2022-02-22” and use this for the following questions.

```{r}
on22 <- filter(sd, as_of_date == "2022-02-22")
```

Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”?

```{r}
mean (on22$percent_of_population_fully_vaccinated, na.rm=TRUE)
```
70.41%

Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22
```{r}
library (ggplot2)

ggplot(on22, aes(x=percent_of_population_fully_vaccinated)) + geom_histogram(binwidth=0.1) + labs(title="Histogram of Vaccination Rates Across San Diego Counties", x="Percent Population Fully Vaccinated", y="Freq")
```


#Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:
```{r}
ucsdplot<- ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(title = "La Jolla Vaccination Rate", x="Date", y="Percent Vaccinated")
ucsdplot
```
Comparing to similar sized areas
Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2022-02-22”.


```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-02-22")

head(vax.36)
```
Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
meanpop<- mean(vax.36$percent_of_population_fully_vaccinated)
ucsdplot + geom_line(y=meanpop)

```
Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?

```{r}
fivenum(vax.36$percent_of_population_fully_vaccinated)
mean (vax.36$percent_of_population_fully_vaccinated)
```


Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36, aes(x=percent_of_population_fully_vaccinated)) + geom_histogram()
```
Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
```{r}
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92040" | zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
Below.

Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.


```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0, 1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate across California",
       subtitle="Only areas with population above 36 K are Shown") +
  geom_hline(yintercept = 0.74, linetype=2)
```

Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

Pretty much okay!



