---
title: "Week05_lab"
author: "Ethan Lai"
output: pdf_document
---

Write a function to grade homework

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

#Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped.


```{r}
# grade() takes one parameter, a numeric vector x.
# grade() excludes

#' Calculates average score for a vector of homework scores, dropping lowest single score. NAs are treated as 0 scores.
#'
#' @param x is a numeric vector of scores
#'
#' @return Average score
#' @export 
#'
#' @examples
#' student <-c(100,NA,90,80)
#' grade(student)
#' 
grade<- function(x){
  
  
  #Convert NA scores to 0
  x[is.na(x)]<-0
  
  #return mean of x with lowest value excluded
  return(mean(x[-which.min(x)]))
}
```
Run Test Cases on individual students:
```{r}
#run Test Cases
grade(student1)
grade(student2)
grade(student3)
```
Now to grade the entire gradebook:
```{r}
url<-"https://tinyurl.com/gradeinput"
gradebook<- read.csv(url,row.names=1)
results<- apply(gradebook, 1, grade)
results
```


Q2. Using your grade() function and the supplied gradebook, who is the top scoring student overall in the gradebook? [3pts]

```{r}
which.max(results)
```
Student 18 is the top scorer.


Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall? [2pts]

```{r}
#Let's try the median score
#Convert NAs to zero
gradebook[is.na(gradebook)]<-0
by_assignment_score<- apply(gradebook,2,median, na.rm=TRUE)
by_assignment_score
which.min(by_assignment_score)
```
HW 2 had the lowest median score.
```{r}
#Let's try the median score
#Convert NAs to zero
gradebook[is.na(gradebook)]<-0
by_assignment_score<- apply(gradebook,2,mean, na.rm=TRUE)
by_assignment_score
which.min(by_assignment_score)
```
HW 2 also had the lowest mean score!


Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
gradebook[is.na(gradebook)]<-0
correlations<- apply(gradebook, 2,cor, x=results)
correlations
which.min(correlations)
```
HW2 is least correlated to overall score.


Q5. Make sure you save your Rmarkdown document and can click the “Knit” button to generate
a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]