---
title: "Class07 Machine Learning 1"
author: 'Ethan Lai'
date: "2/9/0222"
output: pdf_document
---

#Clustering Methods

## K-means

First generate some data to test. Concatenate two rnorm samples, and column bind this vector and its reverse.

```{r}
tmp<- c(rnorm(30,-3), rnorm(30,3))
x<-cbind(x=tmp, y=rev(tmp))
plot(x)
```

```{r}
k <- kmeans(x, centers=2, nstart=10)
k
```

How many points per cluster?
```{r}
k$size
```

What are the centroids of the clusters?
```{r}
k$centers
```

What is the cluster vector?
```{r}
kclusters<- k$cluster
kclusters
```

```{r}
plot(x, col=kclusters)

```
# Recall that R has recycling, so we must be careful
```{r}
y<-1:5
cbind(y,"red")
```

#GG plot version

```{r}
library(ggplot2)
ggplot(data.frame(x),aes(x,y))+geom_point(color=kclusters)
```

#Hierchical Clustering

Must give hclust a distance matrix as in put, not the raw data. hclust objects have a custom plot function, producing a cluster dendrogram.
```{r}
hc<- hclust(dist(x))
plot(hc)
#abline(h=10)

```
Now we "cut" the tree to get our clusters. The function to do so is cutree()

```{r}
ct<- cutree(hc, h=10)

plot(x, col=ct)
```



#  PCA
PCA is a useful method for analyzing large multidimensional datasets

Import data set of UK foods

```{r}

url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url,row.names=1)

```

Let's look at x:

```{r}
head(x)
```

```{r}
nrow(x)
ncol(x)
```


We could plot this with a barplot
```{r}
barplot(as.matrix(x), col=rainbow(nrow(x)))
```
We can unstack with 'besides=TRUE'

```{r}
barplot(as.matrix(x), col=rainbow(nrow(x)), beside=TRUE)

```

One plot that is useful is a pairs plot. This produces a matrix of all pairwise correlations.

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```
# PCA to the rescue

What does PCA tell us about this data?

The main PCA function in base R is called 'prcomp()'

```{r}
# Use the prcomp() PCA function 

# note that prcomp rquires transpose of our matrix
pca <- prcomp( t(x) )
summary(pca)
```

# a plot of PC1 vs PC2 is often called a PCA plot or "score plot"
```{r}
attributes(pca)

```
To generate our score plot, we want pca$x component of the resulting object


```{r}
pca$x

```


```{r}

colorv<- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500), col=colorv, pch=16)
text(pca$x[,1], pca$x[,2], colnames(x), col=colorv)

```
The loadings( aka weights) tell us how the original variables contribute to the PCs

```{r}
barplot(pca$rotation[,1], las=2)
```


#RNA Seq Dataset




