---
title: "CLass10"
author: "Ethan Lai"
date: "2/18/2022"
output: pdf_document
---

What proportion of the MXL sample population is G/G homozygous
```{r}
mxl<- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
table(mxl$Genotype..forward.strand.) /nrow(mxl) *100


```
14.065% are G/G homozygous


#Population Scale Analysis

Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. Hint: The read.table(),
summary() and boxplot() functions will likely be useful here. There is an example R script
online to be used ONLY if you are struggling in vein. Note that you can find the medium value
from saving the output of the boxplot() function to an R object and examining this object.
There is also the medium() and summary() function that you can use to check your
understanding.
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
exprTable<- table(expr$geno)
exprTable
```
108 A/A, 233 A/G, and 121 G/G

To find the medians:
```{r}
exprdf<- data.frame(expr)
exprAA<- exprdf[exprdf[2]=="A/A",]
exprAG<- exprdf[exprdf[2]=="A/G",]
exprGG<- exprdf[exprdf[2]=="G/G",]
median (exprAA[,3])
median (exprAG[,3])
median (exprGG[,3])





```
So the medians for A/A, A/G, and G/G respectively are 31, 25, and 20.

Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
ggplot(expr, aes(x=geno, y=exp)) + geom_boxplot()

```
G/G seems to have the lowest expression, with AA the highest, and heterozygous A/G intermediate between the two homozygous genotypes!


